---
title: "Fit GLMMs to Ascocyta blight spatiotemporal development data"
author: "Ihsan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
knitr::opts_chunk$set(progress = TRUE, verbose = TRUE)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 7,
  fig.height = 7,
  fig.align = "center"
)
```


### Load libraries 

```{r load-libraries, message=FALSE, echo=TRUE, warning=FALSE}
library("readxl")
library("tidyverse")
library("lubridate")
library("spatiotemporaldynamics")
library("SDMTools")
library("glmmTMB")
library("DHARMa")
library("gridExtra")
library("ggpubr")
library("here")
library("ggeffects")
```

### Import disease data

```{r import-diseaes-data}
### import spatiotemporal spread data
dat <-
  read_excel(
    system.file("extdata", "SpatioTemporalSpreadData_N.xlsx",
                package = "spatiotemporaldynamics"),
    sheet = 1
  )

summary(dat)
```

### Import daily average wind direction data

```{r import-wind-direction-data}
wind <- 
  read_excel(
      system.file("extdata", "DailyWindDirectionData.xlsx",
                  package = "spatiotemporaldynamics"),
      sheet = 1
   )

summary(wind)
```

### Convert text to degrees

Wind direction for the Billa Billa site is the character format recorded as the text value of the wind direction.
We need to convert it to degrees for calculations and then calculate the fortnightly average wind direction for use in the GLMMs.

```{r convert-text-to-degrees}
wind <-
  wind %>%
  mutate(wind_degrees = as.numeric(
    case_when(
      daily_avg_wind_direction == "N" ~ "0",
      daily_avg_wind_direction == "NbE" ~ "11.25",
      daily_avg_wind_direction == "NNE" ~ "22.5",
      daily_avg_wind_direction == "NEbN" ~ "33.75",
      daily_avg_wind_direction == "NE" ~ "45",
      daily_avg_wind_direction == "NEbE" ~ "56.25",
      daily_avg_wind_direction == "ENE" ~ "67.5",
      daily_avg_wind_direction == "EbN" ~ "73.5",
      daily_avg_wind_direction == "E" ~ "90",
      daily_avg_wind_direction == "EbS" ~ "101.2",
      daily_avg_wind_direction == "ESE" ~ "112.5",
      daily_avg_wind_direction == "SEbE" ~ "123.8",
      daily_avg_wind_direction == "SE" ~ "135.1",
      daily_avg_wind_direction == "SEbS" ~ "146.3",
      daily_avg_wind_direction == "SSE" ~ "157.6",
      daily_avg_wind_direction == "SbE" ~ "168.8",
      daily_avg_wind_direction == "S" ~ "180",
      daily_avg_wind_direction == "SbW" ~ "191.2",
      daily_avg_wind_direction == "SSW" ~ "202.5",
      daily_avg_wind_direction == "SWbS" ~ "213.8",
      daily_avg_wind_direction == "SW" ~ "225",
      daily_avg_wind_direction == "SWbW" ~ "236.2",
      daily_avg_wind_direction == "WSW" ~ "247.5",
      daily_avg_wind_direction == "WbS" ~ "258.8",
      daily_avg_wind_direction == "W" ~ "270",
      daily_avg_wind_direction == "WbN" ~ "281.2",
      daily_avg_wind_direction == "WNW" ~ "292.5",
      daily_avg_wind_direction == "NWbW" ~ "303.8",
      daily_avg_wind_direction == "NW" ~ "315",
      daily_avg_wind_direction == "NWbN" ~ "326.2",
      daily_avg_wind_direction == "NNW" ~ "337.5",
      daily_avg_wind_direction == "NbW" ~ "348.8",
      TRUE ~   daily_avg_wind_direction
    )
  )) %>%
  group_by(location, assessment_number) %>%
  summarise(wind_direction = circular.averaging(wind_degrees))
```
### Join the wind direction and disease spread data 

Left-join the wind and disease data for analysis in the GLMMs.

```{r join-data}
dat <- left_join(dat, wind, by = c("location", "assessment_number"))
```

### Combine quadrat direction and location columns 

This is required to determine whether quadrats located in particular direction around the primary infection foci has a significantly faster disease progress rates, which in turn will inform us about directional disease spread by wind direction

```{r combine-location-and-quadrat-direction-columns}
dat <-
  unite(dat, quadrat_direc, c(location, direction), remove = FALSE) 
```


### Examine data

```{r examine-data}
str(dat)
```

### Convert data to correct classes

```{r class-coversion}
cols_1 <- c("location", "quadrat", "direction", "quadrat_direc")
dat[cols_1] <- lapply(dat[cols_1], factor)
cols_2 <- c("infected_plants", "total_plants")
dat[cols_2] <- lapply(dat[cols_2], as.integer)
dat$assessment_number <- as.factor(dat$assessment_number)
```


### Re-check class

```{r class-re-check}
sapply(dat, class)
```

### Set seed number for reproducibility purposes

```{r set-seed-number}
set.seed(42)
```

## Fit models

### Mod1 (GLMMs for Tosari & Billa Billa)

Use glmms to include different assessment dates/assessment numbers as random effects since they were not independent. That is, disease assessment conducted at a later assessment date was dependent on the amount of disease present in the former assessment date. 
Density plots show the data is over-dispersed. It should be noted that over-dispersion is due to excess zeros, especially at the beginning, which in turn is attributed to low level of inoculum/rain. That is, excess zeros explains biological phenomena, and we don't want to control for over-dispersion. Rather we are interested in making inferences about over-dispersion as a component of ecological process. The use of quasipoisson family, which fits an extra parameter that allows variance is greater than mean, allows to make such inferences

```{r fit-mod-1}
mod1 <-
   glmmTMB(
      infected_plants ~ total_rain + avg_rh + avg_temp + avg_wind_speed + distance  + location +  (1 |assessment_number) + offset(log(total_plants)),
      family = nbinom1,
      data = dat
   )
                                                                  
summary(mod1)
```

The predictor `quadrat_direc` had to be dropped from the model because the model failed to converge. There is not enough data for the mixed model to include all predictors. 

### Mod2 (Interaction b/w wind speed & direction)

Relative humidity has a negative effect, which doesn't make any biological sense. Try interaction between wind direction and relative humidity 

```{r fit-mod-2, message=FALSE}
mod2 <-
   glmmTMB(
      infected_plants ~ total_rain + avg_rh + wind_direction + avg_rh * wind_direction  + avg_wind_speed + avg_temp + distance + location +
         (1 |  assessment_number) +  offset(log(total_plants)),
      family = nbinom1,
      data = dat
   )
summary(mod2)
```

It can be seen that significant negative effect of relative humidity has been removed with a significant negative interaction between wind speed and wind direction

### Mod3 (Remove wind speed)

Remove wind speed due to its very p-large value

```{r fit-mod-3}
mod3 <-
   glmmTMB(
      infected_plants ~ total_rain + wind_direction + avg_rh * wind_direction +  avg_temp + distance + location + (1 | assessment_number) +  offset(log(total_plants)),
      family = nbinom1,
      data = dat
   )

summary(mod3)
```

## Compare models

### ANOVA
```{r compare-models}
anova(mod1, mod2, mod3)
```

Mod3 explains significant variations. 

## Model diagnostics

Plot best fit model `mod3` 

```{r tosari-billa-billa}
simulateResiduals(mod3, plot = T, quantreg=T)
```

The qq-plot is nearly linear, suggesting that the overall distribution is OK. 

## Marginal effects plots

Remove offset to make prediction for the current number of plants. Retaining them result in very low y-axis values

```{r both-locations, message=FALSE}
both_mod3 <-
   glmmTMB(
      infected_plants ~ total_rain + wind_direction + avg_rh * wind_direction +  avg_temp + distance + location + (1 | assessment_number),
      family = nbinom1,
      data = dat
   )
summary(both_mod3)
```

Make plot for single predictors and later combine them - ggeffects package facet plots by color, which are hard to interpret

```{r plot-predictions}
f1 <- plot(ggpredict(both_mod3 , "total_rain"))
f2 <- plot(ggpredict(both_mod3 , c("avg_rh", "wind_direction")))
f3 <- plot(ggpredict(both_mod3 , "avg_temp"))
f4 <- plot(ggpredict(both_mod3, "distance"))
f5 <- plot(ggpredict(both_mod3, "wind_direction"))
fig_5 <- grid.arrange(f1, f2, f3, f4, f5)
fig_5

ggsave(
   here("man", "figures/fig_5.png"),
   plot = fig_5,
   width = 9,
   height = 9,
   units = "in",
   dpi = 600
)

fig_5

dev.off()
```

